---
title: "R notes base R"
author: "Y.Fan"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### bracket
- "()" is used with functions
- "[]" is for indexing
- "[[]]" is to call the elements in a list
- "{}" is to define a block of code within a function

### Missing values
is.na() gives a vector of logical\
anyNA() gives one logical
```{r na}
mean(c(1,2,NA), na.rm = TRUE)
is.na(c(1,2,NA,NA,5))
anyNA(c(1,2,NA,3))
```

### Factor
The fist argument of factor() must be 'x=yourvector'.
```{r factor}
factor(x = c('A','B','C'), levels = c('A','C'), ordered = TRUE)
```

### Sorting vectors
sort() by default sort in increasing order\
use argument `decreasing = TRUE` for decreasing order
```{r}
sort(c(1,3,4,2), decreasing = TRUE)
```

#### exercise sorting data frames
Load the airquality data set.
```{r}
data("airquality")
```

Open the help page for this data set.
```{r}
help("airquality")
```

Examine the data set.
```{r}
summary(airquality)
```

Display the column names of the airquality data frame.
```{r}
colnames(airquality)
```

Sort the data frame by increasing value in the Ozone column.
```{r}
o <- order(airquality$Ozone)
airquality[o,]
```

Sort the data frame by Month in increasing order and Temp in decreasing order.\
Use the "-" to indicate decreasing can only work with numeric type of data.
```{r}
o <- order(airquality$Month, -airquality$Temp)
airquality[o,]
```

Write the latest sorted data frame to a text file format of your choice.
```{r}
write.table(airquality, file = "sorted.tsv", sep = "\t", row.names = FALSE)
```

#### exercise merging data frames
```{r}
buildings <- data.frame(site = c(1, 2, 3),name = c("b1","b2","b3"))
survey_data <- data.frame(survey = c("A","A","A","B","B","B"),
                          location = c(1,2,3,2,3,1),
                          efficiency = c(51,64,70,71,80,58))
buildings_survey <- merge(x = buildings, y = survey_data, by.x = "site", by.y = "location")
buildings_survey
```

#### exercise summarising groups of data
Compute the mean of each numeric column each month in the airquality data frame using aggregate().
Make sure NA values are removed.
```{r}
summary(airquality)
aggregate(x = airquality, by = list(month = airquality$Month), FUN = mean, na.rm = TRUE)
```

Compute the mean of the Solar.R column each month.
Make sure the grouping column is called Month in the return value.
Make sure NA values are removed.
```{r}
aggregate(x = airquality[,c("Solar.R","Temp")], by = list(month = airquality$Month), FUN = mean, na.rm = TRUE)
```

### Functions
The ellipsis
- a place holder to pass any thing beyond definition of the function to any downstream methods
The default value
- can be set in the "functionname <- function(a, b = 0, c = 0, ...) {}"

#### excercise
Write a function to calculate the hypotenuse of a triangle given the length of the other two sides.
Run the function you have created with diﬀerent values.
```{r}
hypotenuce <- function(a, b, ...){
  message("calculating the hypotenuce of ",a," and ",b)
  c <- sqrt(a^2 + b^2)
  return (c)
}
hypotenuce(3,4)
```

### apply functions
#### exercise
Part 1
Create a vector of integers from 1 to 10.
Compute the log2 of each value in the vector using either lapply() or sapply().
```{r}
a <- c(1L:10L)
# a:b is integer type
a <- 1:10 
lapply(a, FUN = log)
sapply(a, FUN = log)
```

Part 2
Create a list of four elements, each element being a vector of type either numeric or logical.
Compute the sum of each vector in the list using either lapply() or sapply().
```{r}
b <- list(c(1,2,3,4),c(TRUE, FALSE, FALSE), c(1.1,2.2,3.3,4.4), c(TRUE, TRUE))
lapply(b, FUN = sum)
```

Part 3
Use sapply() on the list that you created in part 2, to repeat each element of each vector three times.
i.e., 1, 2, 3 should become 1, 1, 1, 2, 2, 2, 3, 3, 3
```{r}
# help on the rep shows that it can take ..., so put the arguemnt for rep directly in the following
sapply(b, FUN = rep, each = 3)
```

### loops and conditions
#### exercises
Write a for loop that iterates over the integers 1 to 7 and prints the number raised to the power of three.
```{r}
for (i in 1:7){
  print(i^3)
}
```

Write a for loop that iterates over the names of the columns in the builtin data set iris and prints each
column name together with the number of characters in that column name.
- don't expect to write a loop perfectly for the first time
```{r}
data("iris")
for (i in 1:ncol(iris)){
  n <- colnames(iris)
  print(paste0(n[i],", number of characters:", nchar(n[i])))
}
# "i in some_vector" >>>
# for (i in colnames(iris)){
# }
```

Use the ifelse() function to print the name of colours that are made up of four characters in the vector
my_colours below.
```{r}
my_colours <- c("red","orange","purple","yellow","pink","blue")
ifelse(nchar(my_colours) == 4, my_colours, "")
```














